[Artifact Presentation] Damegender: \\ Writing and Comparing Gender Detection Tools

### 2020-05-15 paper2sattose.tex
Version submitted

\begin{abstract}
%Context
Diversity in software development teams have been identified as one of the main ingredients of a more productive, more healthy software community.
Thus, the interest of the research community in identifying who is contributing has increased in the last years.
In the software domain, and although other types of diversity exist, this is especially important for the case of gender.
% Objective, 
Given the large amount of publicly available data on the software development process that can be retrieved and analyzed from the Internet (e.g., GitHub, StackOverflow), the importance of having methods and tools that help with large amounts of data would be desirable.
% Method
In this paper we present a free software tool, called \texttt{damegender}, which we have conceived to given a name outputs the gender and a probability.
\texttt{damegender} is based on open databases from official census and uses Machine Learning to guess strings not classified as names, such as diminutives or nicknames.
% Results 
We have compared \texttt{damegender} with other tools, obtaining good results.
% Conclusions.

\end{abstract}

@@

In Section 2 we explain \texttt{damegender} as solution to the
problem of gender guessing. Section 3 introduces the Open Datasets used as the \emph{golden truth}.
Section 4 presents a feature comparison with other gender guessing tools and services.
Section 5 reports on values of accuracy and offers a confusion matrix using a scientific dataset.
Section 6 is about how we use Machine Learning in \texttt{damegender}.
Section 7 discusses limitations and further research, and concludes the paper.

@@

Gender guesser tools apply several methods for estimating the gender from a given name. 
As a starting point, however, all of them rely on a dataset that contains information on what gender a name usually can be attributed to.

@@

\begin{table*}[ht]
\footnotesize
\begin{tabular}[]{lcccccc}
\hline
Service / Tool $->$ & Gender API & gender-guesser & genderize.io & NameAPI & NamSor & damegender\tabularnewline
\hline
Database size & 431M & 45K & 114M & 1M & 4G & 57K \tabularnewline
Regular data updates & Yes & no & No & Yes & Yes & Yes\tabularnewline
Unstructured full name strings & Yes & No & No & Yes & No & Yes\tabularnewline
Surnames & Yes & No & No & Yes & Yes & Yes\tabularnewline
Non-Latin alphabets & Partial & No & Partial & Yes & Yes & No\tabularnewline
Implicit geo-localization & Yes & No & No & Yes & Yes & No\tabularnewline
Exists locale & Yes & Yes & Yes & Yes & Yes & Yes\tabularnewline
Assignment type & P & B & P & P & P & P \tabularnewline
Free parameters & T,P & G & P,C & T & S & T,C\tabularnewline
Prediction & No & No & No & No & No & Yes\tabularnewline
Free license & No & Yes & No & No & No & Yes\tabularnewline
REST API & Yes & No & Yes & Yes & Yes & Planned\tabularnewline
Limits number of requests & Yes (200) & \infinity & Yes & Yes & Yes & \infinity \tabularnewline
Subscription (100K requests/month)	 & 79 & 0 & 7 & 150 & 80 & 0 \tabularnewline
\hline
\end{tabular}
\caption{Comparison of the different features that gender guesser software services and tools offer. Assignment type = \{P: Probabilistic; B: Binary\}. Free Parameters = \{T: total\_names; P: probability; C: count; G: gender; T: trust; S: scale \}. The subscription price is given in euro.}
\label{table:comparison}
\end{table*}

@@

The third approach is based on the trust on commercial solutions, in the same way we trust search engines when we make searches in Internet.
This is because commercial APIs can be seen just a black box, so we do not know where the data comes from and how it has been treated.
As at this point, commercial APIs offer better results as other solutions, \texttt{damegender} gives the possibility to include data from them.
Thus, it is possible to download JSON files from the main commercial gender guesser API solutions (e.g., \texttt{genderapi}, \texttt{genderize}, \texttt{namsor}, \texttt{nameapi}) and use it as the dataset.
There are certain uses that are currently only available in such tools.

@@

The market of gender detection tools and services is currently dominated by companies based on payment services through Application Programming Interfaces.
Without doubt, they offer good results, with high accuracy values.
However, their inner working cannot be studied (i.e., they work as a black-box for the outsider) and the fees that have to be paid for using their service are sometimes out of the reach of many researchers.
That is why we propose a new tool, called \texttt{damegender}, with the aim of having open data on gender guessing, which offers more flexibility and where researchers can build on top of it.
This tool is offered under a free software license and is available on GitHub for download and enhancement.
As we have shown in this paper, although still incipient, the tool offers good accuracy values based on the use of public databases from government bodies and on the use of machine learning algorithms.
Nonetheless, we have to note that \texttt{damegender} is still under development, and that it has to be applied to several real repositories to confirm its benefits and address its limitations (such as a small database size of gendered names).

@@

We would like to thank: Lucía Santamaría and Helena Mihaljevic by the previous work; Daniel Izquierdo and Laura Arjona for starting this research field at URJC; Jesus González Boticario and the people who work with him at UNED for the motivation towards the machine learning; all people working with Jesus González Barahona and Gregorio Robles.

>>>>

\begin{abstract}
%Context
Diversity in software development teams have been identified as one of the main ingredients of a more productive, more healthy software community.
Thus, the interest of the research community in identifying who is contributing has increased in the last years.
In the software domain, and although other types of diversity exist, this is especially important for the case of gender.
% Objective, 
Given the large amount of publicly available data on the software development process that can be retrieved and analyzed from the Internet (e.g., GitHub, StackOverflow), the importance of having methods and tools that help with large amounts of data would be desirable.
% Method
In this paper we present a free software tool, called \texttt{damegender}, which we have conceived to given a name outputs the gender and a probability.
\texttt{damegender} is based on open databases from official census and uses Machine Learning to guess strings not classified as names, such as diminutives or nicknames.
% Results 
We have compared \texttt{damegender} with other tools, obtaining good results.
% Conclusions.
We hope the \texttt{damegender} tool can become a cornerstone for the scientific advancement of the study of gender (including gender gap) and diversity in the IT, and in particular in the free software community.
\end{abstract}

@@

In Section 2 we explain \texttt{damegender} as solution to the
problem of name-to-gender inference. Section 3 introduces the Open Datasets used as the \emph{golden truth}.
Section 4 presents a feature comparison with other name-to-gender inference services and tools.
Section 5 reports on values of accuracy and offers a confusion matrix using a scientific dataset.
Section 6 is about how we use Machine Learning in \texttt{damegender}.
Section 7 discusses limitations and further research, and concludes the paper.

@@

Name-to-gender inference services and tools apply several methods for estimating the gender from a given name. 
As a starting point, however, all of them rely on a dataset that contains information on what gender a name usually can be attributed to.

@@

\begin{table*}[ht]
\footnotesize
\begin{tabular}[]{lcccccc}
\hline
Service / Tool $->$ & Gender API & gender-guesser & genderize.io & NameAPI & NamSor & damegender\tabularnewline
\hline
Database size & 431M & 45K & 114M & 1M & 4G & 57K \tabularnewline
Regular data updates & Yes & no & No & Yes & Yes & Yes\tabularnewline
Unstructured full name strings & Yes & No & No & Yes & No & Yes\tabularnewline
Surnames & Yes & No & No & Yes & Yes & Yes\tabularnewline
Non-Latin alphabets & Partial & No & Partial & Yes & Yes & No\tabularnewline
Implicit geo-localization & Yes & No & No & Yes & Yes & No\tabularnewline
Exists locale & Yes & Yes & Yes & Yes & Yes & Yes\tabularnewline
Assignment type & P & B & P & P & P & P \tabularnewline
Free parameters & T,P & G & P,C & T & S & T,C\tabularnewline
Prediction & No & No & No & No & No & Yes\tabularnewline
Free license & No & Yes & No & No & No & Yes\tabularnewline
REST API & Yes & No & Yes & Yes & Yes & Planned\tabularnewline
Limits number of requests & Yes (200) & \infinity & Yes & Yes & Yes & \infinity \tabularnewline
Subscription (100K requests/month)	 & 79 & 0 & 7 & 150 & 80 & 0 \tabularnewline
\hline
\end{tabular}
\caption{Comparison of the different features that name-to-gender inference services and tools offer. Assignment type = \{P: Probabilistic; B: Binary\}. Free Parameters = \{T: total\_names; P: probability; C: count; G: gender; T: trust; S: scale \}. The subscription price is given in euro.}
\label{table:comparison}
\end{table*}

@@

The third approach is based on the trust on commercial solutions, in the same way we trust search engines when we make searches in Internet.
This is because commercial APIs can be seen just a black box, so we do not know where the data comes from and how it has been treated.
As at this point, commercial APIs offer better results as other solutions, \texttt{damegender} gives the possibility to include data from them.
Thus, it is possible to download JSON files from the main commercial name-to-gender inference API solutions (e.g., \texttt{genderapi}, \texttt{genderize}, \texttt{namsor}, \texttt{nameapi}) and use it as the dataset.
There are certain uses that are currently only available in such tools.

@@

The market of gender detection tools and services is currently dominated by companies based on payment services through Application Programming Interfaces.
Without doubt, they offer good results, with high accuracy values.
However, their inner working cannot be studied (i.e., they work as a black-box for the outsider) and the fees that have to be paid for using their service are sometimes out of the reach of many researchers.
That is why we propose a new tool, called \texttt{damegender}, with the aim of having open data on name-to-gender inference, which offers more flexibility and where researchers can build on top of it.
This tool is offered under a free software license and is available on GitHub for download and enhancement.
As we have shown in this paper, although still incipient, the tool offers good accuracy values based on the use of public databases from government bodies and on the use of machine learning algorithms.
Nonetheless, we have to note that \texttt{damegender} is still under development, and that it has to be applied to several real repositories to confirm its benefits and address its limitations (such as a small database size of gendered names).

@@

We would like to thank: Lucía Santamaría and Helena Mihaljevic for the previous work; Daniel Izquierdo and Laura Arjona for starting this research field at URJC; Jesus González Boticario and the people who work with him at UNED for the motivation towards the machine learning; all people working with Jesus González Barahona and Gregorio Robles.
